---
title: "Heart Disease Analysis and Rediction"
author: "Chia-Yu Wei"
date: "2024-04-18"
output:
  word_document: default
  pdf_document: default
subtitle: Multivariate Analysis Final Project
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(car)
library(MASS)
library(lattice)
library(ggplot2)
library(patchwork)
library(factoextra)
```

# 1    Introduction

  According to the CDC, heart disease is the leading cause of death across gender and racial/ethnic groups in United State, with one person dying every 33 seconds from cardiovascular disease. Heart disease can have significant impacts on only family but country. Early diagnosis and treatment of heart disease can significantly reduce the occurrence and severity of the disease. The question is “can we predict if the patients will suffer from heart disease through some features”.

## Data Overview
```{r, echo=FALSE}
data <- read.csv("heart.csv")
data <- data[, -c(3, 7)]
```

### Continuous Variables

I noticed that there are so many zero in *Cholesterol*, but it's possible to have value of cholesterol is 0. They might be missing data and affect the model a lot. Therefore, I removed them from the data set.
```{r, echo=FALSE}
ggplot(data = data, aes(x = Age, fill = as.factor(HeartDisease))) + 
  geom_density(alpha = 0.5) +  
  labs(x = "Age", y = "Density", fill = "Heart Disease") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red")) +
  theme_minimal()
ggplot(data = data, aes(x = RestingBP, fill = as.factor(HeartDisease))) + 
  geom_density(alpha = 0.5) +  
  labs(x = "RestingBP", y = "Density", fill = "Heart Disease") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"))  +
  theme_minimal()
ggplot(data = data, aes(x = Cholesterol, fill = as.factor(HeartDisease))) + 
  geom_density(alpha = 0.5) +  
  labs(x = "Cholesterol", y = "Density", fill = "Heart Disease") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"))  +
  theme_minimal()
ggplot(data = data, aes(x = MaxHR, fill = as.factor(HeartDisease))) + 
  geom_density(alpha = 0.5) +  
  labs(x = "MaxHR", y = "Density", fill = "Heart Disease") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"))  +
  theme_minimal()
ggplot(data = data, aes(x = Oldpeak, fill = as.factor(HeartDisease))) + 
  geom_density(alpha = 0.5) +  
  labs(x = "Oldpeak", y = "Density", fill = "Heart Disease") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"))  +
  theme_minimal()
```

### Discrete Variables
```{r}
zero <- data[, 4] == 0
#n = 746
data <- data[!zero, ]
ggplot(data = data, aes(x = Sex, fill = as.character(HeartDisease))) +
  geom_bar() +
  labs(title = 'Sex and Heart Diagnoses',
       x = 'Sex',
       y = 'Frequency')
ggplot(data = data, aes(x = FastingBS, fill = as.character(HeartDisease))) +
  geom_bar() +
  labs(title = 'FastingBS and Heart Diagnoses',
       x = 'FastingBS',
       y = 'Frequency')
ggplot(data = data, aes(x = ExerciseAngina, fill = as.character(HeartDisease))) +
  geom_bar() +
  labs(title = 'ExerciseAngina and Heart Diagnoses',
       x = 'ExerciseAngina',
       y = 'Frequency')
ggplot(data = data, aes(x = ST_Slope, fill = as.character(HeartDisease))) +
  geom_bar() +
  labs(title = 'Slope and Heart Diagnoses',
       x = 'ST_Slope',
       y = 'Frequency')
```

## 2    Logisitc Regression
```{r, echo=FALSE, results='hide'}
data$ST_Slope <- factor(data$ST_Slope, levels = c("Up", "Flat", "Down"))
model <- glm(HeartDisease ~ ., data = data, family = binomial)
summary(model)
```

### Model selection
```{r, echo=FALSE}
null_model <- glm(HeartDisease ~ 1, data = data, family = binomial)
step_model <- stepAIC(null_model, direction = "forward",
                      scope = ~ Age + Sex + RestingBP + Cholesterol + FastingBS + MaxHR + ExerciseAngina + Oldpeak + ST_Slope)
step_model$coefficients
summary(step_model)
```

###Diagnose

```{r, echo=FALSE}
final_model <- glm(HeartDisease ~. - RestingBP - MaxHR - FastingBS, data = data, family = binomial)
vif(final_model, echo=FALSE)
```

### Prediction
```{r, echo=FALSE}
library(caret)
split <- createDataPartition(data$HeartDisease, p = 0.80, list = FALSE)
train_data <- data[split, ]
test_data <- data[-split, ]
model <- glm(HeartDisease ~ ., data = train_data, family = binomial())
predictions <- predict(step_model, test_data, type = "response")
predicted_classes <- ifelse(predictions > 0.5, 1, 0)
confusionMatrix <- table(test_data$HeartDisease, predicted_classes)
print(confusionMatrix)
accuracy <- sum(diag(confusionMatrix)) / sum(confusionMatrix)
print(paste("Accuracy:", accuracy))
```
```{r}
library(pROC)
test_prob = predict(final_model, newdata = test_data, type = "response")
par(pty = "s")
test_roc = roc(test_data$HeartDisease ~ test_prob, plot = TRUE,
               print.auc = TRUE,legacy.axes=TRUE)
```

### PCA
```{r}
data_num <- data[, c(1, 3, 4, 6, 8)]
data_num_scaled <- scale(data_num)
head(data_num_scaled)
pca <- prcomp(data_num_scaled, center = TRUE, scale. = TRUE)
summary(pca)

fviz_eig(pca, addlabels = TRUE)

pc_scores <- pca$x
```

### Canonical Correlation Analysis
```{r, echo=FALSE}
data$ST_Slope <- ifelse(data$ST_Slope == "Up", 1, ifelse(data$ST_Slope == "Flat", 2, 3))
data$ExerciseAngina <- ifelse(data$ExerciseAngina == "N", 0, 1)
data$FastingBS <- as.numeric(data$FastingBS)
data$Sex <- ifelse(data$Sex == "M", 0, 1)
demo_set <- data[, c(1, 2)]
rest_set <- data[, c(3, 4, 5, 6)]
heart_set <- data[, c(7, 8, 9)]
Y_set <- as.numeric(data[, 10])
(cca_result1 <- cancor(demo_set, rest_set))
(cca_result2 <- cancor(rest_set, heart_set))
(cca_result3 <- cancor(demo_set, heart_set))
(cca_result4 <- cancor(demo_set, Y_set))
cancor(rest_set, Y_set)
cancor(heart_set, Y_set)
```
```{r, echo=FALSE}
library(reshape2)
library(ggplot2)
data$HeartDisease <- as.numeric(data$HeartDisease)
data_norm <- scale(data)
cor_matrix <- cor(data_norm)

melted_cor_matrix <- melt(cor_matrix)


ggplot(data = melted_cor_matrix, aes(Var1, Var2, fill = value)) +
    geom_tile() + 
    geom_text(aes(label = round(value, 2)), vjust = 1, size = 3) +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                         midpoint = 0, limit = c(-1,1), space = "Lab", 
                         name="Pearson\nCorrelation") +
    theme_minimal() +  
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), 
          axis.title = element_blank()) + 
    coord_fixed()

```

```{r, echo=FALSE}
data_scaled <- scale(data)
# K-means clustering
k <- 2  # assuming the elbow plot showed 3 as a good number
kmeans_result <- kmeans(data_scaled, centers = k, nstart = 10)
fviz_cluster(kmeans_result, data = data_scaled, geom = "point",
             xlab = "PC1", ylab = "PC2")
```

```{r}
heart <- ifelse(data$HeartDisease == 0, 1, 2)
cm <- table(heart, kmeans_result$cluster) 
cm
(accuracy <- (cm[1, 1] + cm[2, 2])/sum(cm))
```
```{r}
new_data_lb <- ifelse(step_model$fitted.values < 1/3, 1,
                   ifelse(step_model$fitted.values < 2/3, 2, 3))
new_data <- data
new_data$HeartDisease <- new_data_lb
data_scaled <- scale(new_data)

wss <- sapply(1:10, function(k){kmeans(data_scaled, k, nstart = 10)$tot.withinss})
plot(1:10, wss, type = "b", xlab = "Number of Clusters", ylab = "Within groups sum of squares")

k <- 3  # assuming the elbow plot showed 3 as a good number
kmeans_result_k3 <- kmeans(data_scaled, centers = k, nstart = 10)
fviz_cluster(kmeans_result_k3, data = data_scaled, geom = "point",
             xlab = "PC1", ylab = "PC2")
```

```{r}
library(e1071)
#tune.svm.linear <- tune.svm(HeartDisease ~ ., data = train_data, kernel = "linear" , cost=c(0.001,0.01,0.1), gamma=c(0.001,0.01,0.1))

tune.svm.radial <- tune.svm(HeartDisease ~ ., data = train_data, kernel = "radial" , cost=c(0.01,0.1,0.5,1), gamma=c(0.1,0.5,1))

#summary(tune.svm.linear)
```

```{r}
#svm.linear <- svm(HeartDisease ~ ., data = train_data, kernel = "linear", cost = 0.01, gamma = 0.001)
svm.radial <- svm(HeartDisease ~ ., data = train_data, kernel = "radial", cost = 1, gamma = 0.1)

#svm.linear.pred <- predict(svm.linear, newdata = test_data)
#svm.linear.pred <- ifelse(svm.linear.pred >= 0.5, 1, 0)
#table(test_data$HeartDisease,svm.linear.pred)
#sum(test_data$HeartDisease == svm.linear.pred)/nrow(test_data)

svm.radial.pred <- predict(svm.radial, newdata = test_data)
svm.radial.pred <- ifelse(svm.radial.pred >= 0.5, 1, 0)
table(test_data$HeartDisease,svm.radial.pred)
sum(test_data$HeartDisease == svm.radial.pred)/nrow(test_data)
```